Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)

### 2025-05-21 05-discussion.qmd
ajustes

The third hypothesis concerned the moderating role of perceptions of inequality on the relationship between meritocracy and market justice attitudes. The analysis indicated that the positive association between meritocratic beliefs and support for market-based distribution tended to weaken when perceptions of economic inequality were particularly high. This pattern suggested that high awareness of disparities could challenge the moral high ground of meritocracy, as individuals might question whether success was truly based on effort or whether systemic biases distorted outcomes. Although the moderation was not consistently statistically significant across all models (i.e.within subjects effects), the observed trend aligned with theoretical expectations that perceptions of inequality undermined the legitimacy of meritocratic narratives and fragmented support for existing economic arrangements. This interaction revealed the complex psychological landscape in which deeply held beliefs about effort and reward could be destabilized by heightened awareness of systemic disparities, especially during times of social upheaval or mobilization.

>>>>

The third hypothesis concerned the moderating role of perceptions of inequality on the relationship between meritocracy and market justice attitudes. The analysis indicated that the positive association between meritocratic beliefs and support for market-based distribution tended to weaken when perceptions of economic inequality were particularly high. Although this moderation is not significant for within-subject effects (i.e. over time), when taking into account between-subjects differences we found a positive effect. This appears as contrary to our initial hypothesis, as it would mean that effect of meritocratic perceptions becomes stronger as perception of inequality increases. Nevertheless, we need to be cautious about this interpretation, as the main effect of inequality perception on market justice is negative. Therefore, as perceived inequality become less negative (i.e. more positive), or closer to zero, then the meritocratic perception has an stronger positive relationship on market justice preferences. This could mean that meritocratic perceptions play an stronger role in justification of market justice inequalities for those who are less capable or less prone to perceive inequality.

### 2025-05-21 06-conclusion.qmd
ajustes

The findings confirm that higher perceptions of economic inequality are associated with increased support for market justice attitudes, this is, the belief that it is fine that those with higher income have better social services such as education, pensions and health. Meritocratic beliefs strongly reinforce support for market justice, serving as moral justifications for structural inequalities. Additionally, perceptions of inequality moderate the influence of meritocratic beliefs, such that high awareness of disparities can weaken support for existing market-based allocations. Notably, the social mobilizations of 2019 and subsequent periods seem to have contributed to shifting attitudes, fostering greater skepticism of systemic fairness and reducing the legitimizing power of meritocratic narratives.

@@

Despite its contributions, the study has several limitations. First, the reliance on self-reported perceptions and attitudes may be subject to social desirability bias. Second, the focus on Chile, while offering valuable insights into a specific context, limits the generalizability of results to other countries with different institutional and cultural backgrounds. Third, the observational nature of the data prevents causal inferences, and unobserved confounding factors may influence the observed relationships. Finally, the potential influence of media, political discourse, and peer effects on perceptions was not directly assessed.

Future research could explore the causal mechanisms linking perceptions and attitudes, possibly through experimental or mixed-method designs. Comparative studies across different national contexts would help determine the universality or specificity of these dynamics. Additionally, investigating the role of media, political communication, and education in shaping perceptions of inequality and meritocracy would deepen understanding of the normative foundations of social justice attitudes. Finally, examining how these perceptions influence behavioral outcomes, such as political participation or support for social movements, would provide valuable insights into the pathways from beliefs to collective action and policy change.

>>>>

The findings confirm that higher perceptions of economic inequality are associated with less support for market justice attitudes, this is, the belief that it is fine that those with higher income have better social services such as education, pensions and health. Meritocratic beliefs strongly reinforce support for market justice, serving as moral justifications for structural inequalities. Additionally, perceptions of inequality moderate the influence of meritocratic beliefs, such that as the awareness of disparities are attenuated (i.e. less negative) the role of meritocratic beliefs on existing market-based allocations become stronger. Notably, the social mobilizations of 2019 and subsequent periods seem to have contributed to shifting attitudes, fostering greater skepticism of systemic fairness and reducing the legitimizing power of meritocratic narratives.

@@

Despite its contributions, the study has several limitations. First, the reliance on self-reported perceptions and attitudes may be subject to social desirability bias. Second, the focus on Chile, while offering valuable insights into a specific context, limits the generalizability of results to other countries with different institutional and cultural backgrounds. Third, the observational nature of the data prevents causal inferences, and unobserved confounding factors may influence the observed relationships. Finally, the potential influence of media, political discourse, and peer effects on perceptions was not directly assessed. Future research could explore the causal mechanisms linking perceptions and attitudes, possibly through experimental or mixed-method designs. Comparative studies across different national contexts would help determine the universality or specificity of these dynamics. Additionally, investigating the role of media, political communication, and education in shaping perceptions of inequality and meritocracy would deepen understanding of the normative foundations of social justice attitudes. Finally, examining how these perceptions influence behavioral outcomes, such as political participation or support for social movements, would provide valuable insights into the pathways from beliefs to collective action and policy change.
